Sparse Function-space Representation of Neural Networks

### 2023-08-31 workshops/icml-duality/main.tex
updated

\begin{figure}[t!]
  % \centering\scriptsize
  \centering\Large
  % Figure options
  \pgfplotsset{axis on top,scale only axis,width=\figurewidth,height=\figureheight, ylabel near ticks,ylabel style={yshift=-2pt},y tick label style={rotate=90},legend style={nodes={scale=1., transform shape}},tick label style={font=\Large,scale=1}}
  \pgfplotsset{xlabel={Input, $x$},axis line style={rounded corners=2pt}}
  % Set figure
  \setlength{\figurewidth}{.41\textwidth}
  \setlength{\figureheight}{\figurewidth}
  %
  \def\inducing{\Huge Sparse inducing points}
  %
  \begin{subfigure}[c]{.52\columnwidth}
    \raggedleft
    \pgfplotsset{ylabel={Output, $y$}}
    \input{./fig/regression-nn.tex}%
  \end{subfigure}
  \hfill
  \begin{subfigure}[c]{.02\columnwidth}
    \centering
    \tikz[overlay,remember picture]\node(p0){};
  \end{subfigure}
  \hfill
  \begin{subfigure}[c]{.4\columnwidth}
    \raggedleft
    \pgfplotsset{yticklabels={,,},ytick={\empty}}
    \input{./fig/regression-nn2svgp.tex}%
  \end{subfigure}
%  \hfill
%  \begin{subfigure}[c]{.01\textwidth}
%    \centering
%    \tikz[overlay,remember picture]\node(p1){};
%  \end{subfigure}
%  \hfill
%  \begin{subfigure}[c]{.28\textwidth}
%    \raggedleft
%    \pgfplotsset{yticklabels={,,},ytick={\empty}}
%    \input{./fig/regression-update.tex}%
%  \end{subfigure}
  \vspace*{-0.8em}
  \caption{\textbf{Regression example on an MLP with two hidden layers.} Left:~Predictions from the trained neural network. Right:~Our approach (\our) equips trained NNs with uncertainty estimates.} %The model captures the predictive mean and uncertainty, and (right) makes it possible to incorporate new data without retraining the model.}
  % \caption{\textbf{Regression example on an MLP with two hidden layers.} Left:~Predictions from the trained neural network. Right:~Our approach (\our) summarizes all the training data with the help of a set of inducing points.} %The model captures the predictive mean and uncertainty, and (right) makes it possible to incorporate new data without retraining the model.}
  \label{fig:teaser}
  %
  \begin{tikzpicture}[remember picture,overlay]
    % Arrow style
    \tikzstyle{myarrow} = [draw=black!80, single arrow, minimum height=14mm, minimum width=2pt, single arrow head extend=4pt, fill=black!80, anchor=center, rotate=0, inner sep=5pt, rounded corners=1pt]
    % Arrows
    \node[myarrow] (p0-arr) at ($(p0) + (0.2em,1.2em)$) {};
    % \node[myarrow] (p0-arr) at ($(p0) + (1em,1.5em)$) {};
    %\node[myarrow] (p1-arr) at ($(p1) + (1em,1.5em)$) {};
    % Arrow labels
    \node[font=\scriptsize\sc,color=white] at (p0-arr) {\our};
    %\node[font=\scriptsize\sc,color=white] at (p1-arr) {new data};
  \end{tikzpicture}
  \vspace*{-1em}
\end{figure}

@@

\begin{figure*}[t]
  \centering\scriptsize
  % \setlength{\figurewidth}{.18\textwidth}
  \setlength{\figurewidth}{.27\textwidth}
  \setlength{\figureheight}{\figurewidth}
  \pgfplotsset{axis on top,scale only axis,y tick label style={rotate=90}, x tick label style={font=\footnotesize},y tick label style={font=\footnotesize},title style={yshift=-4pt,font=\large}, y label style={font=\large},x label style={font=\large},grid=major, width=\figurewidth, height=\figureheight}
  \pgfplotsset{grid style={line width=.1pt, draw=gray!10,dashed}}
  \pgfplotsset{xlabel={$M$ as \% of $N$},ylabel style={yshift=-12pt}}
  %
  \begin{minipage}[t]{.155\textwidth}
    \raggedleft
    \pgfplotsset{ylabel=NLPD}
    \input{./fig/australian.tex}
  \end{minipage}
  \hfill
  % \begin{minipage}[t]{.12\textwidth}
  %   \raggedleft
  %   \input{./fig/breast_cancer.tex}
  % \end{minipage}
  % \hfill
  % \begin{minipage}[t]{.12\textwidth}
  %   \raggedleft
  %   \input{./fig/ionosphere.tex}
  % \end{minipage}
  % \hfill
  \begin{minipage}[t]{.155\textwidth}
    \raggedleft
    \input{./fig/glass.tex}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{.155\textwidth}
    \raggedleft
    \input{./fig/vehicle.tex}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{.155\textwidth}
    \raggedleft
    \input{./fig/waveform.tex}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{.155\textwidth}
    \raggedleft
    \input{./fig/digits.tex}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{.155\textwidth}
    \raggedleft
    \input{./fig/satellite.tex}
  \end{minipage}\\[-1em]
  %
  % Legend
  \definecolor{steelblue31119180}{RGB}{31,119,180}
  \definecolor{darkorange25512714}{RGB}{255,127,14}
  \newcommand{\myline}[1]{\protect\tikz[baseline=-.5ex,line width=1.6pt]\protect\draw[draw=#1](0,0)--(1.2em,0);}
  \caption{Comparison of convergence in terms of number of inducing points $M$ in NLPD (mean$\pm$std over 5 seeds) on UCI classification tasks: \our (\myline{steelblue31119180}) vs. GP subset (\myline{darkorange25512714}). Our \our converges fast for all cases showing clear benefits of its ability to summarize all the data onto a sparse set of inducing points. The number of inducing points $M$ are specified as a percent of the number of training data $N$.}
  \label{fig:uci}
  %\vspace*{-6pt}
\end{figure*}

@@

As demonstrated in \citet{adam2021dual}, the posterior under this model bears a structure akin to \cref{eq:gp_pred}. The authors of that paper exploit the dual parameters for approximate inference, but write them using the natural parameterization, primarily because this form is more suitable for optimization through natural gradients.
We project the dual parameters onto the inducing points sparse GP using the dual parameters,
% In order to link the dual parameters defined in \cref{eq:dual_param}, we write them in the sparse GP using the dual parameters,
%
\begin{equation} \textstyle
  \valpha_{\vu}  =  \sum_{i=1}^N  \vkzi \, \hat{\alpha}_{i}
  \quad \text{and} \quad
  \MBeta_{\vu} =  \sum_{i=1}^N \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,
\label{eq:dual_sparse}
\end{equation}
%
where the sparse dual parameters are now a sum over \emph{all data points}, with $\valpha_{\vu} \in \R^{M}$ and $\MBeta_{\vu} \in \R^{M  \times M}$. Using this sparse definition of the dual parameters, our sparse GP posterior takes the following form:
\begin{align}\label{eq:dual_sparse_post}
   \myexpect_{q_{\vu}(\vf)}[f_i] &= \vkzs^{\T} \MKzz^{-1} \valpha_{\vu}, \\
   %\quad \text{and} \quad
   \textrm{Var}_{q_{\vu}(\vf)}[f_i]  &= k_{ii} - \vkzs^\top [\MKzz^{-1} - (\MKzz + \MBeta_{\vu})^{-1} ]\vkzs, \nonumber
\end{align}
where $\MKzz$ and $\vkzs$ are defined similarly to $\MKxx$ and $\vk_{\vx i}$ but over the inducing points $\{\vz_j\}_{i=1}^M$, $\vz \in \R^{D}$, instead of the full data set $\dataset$. The key quantities we need to make predictions from our sparse GP from the converged NN are $(\hat{\alpha}_i, \hat{\beta}_i)$ (\cref{eq:dual_param_laplace}) and a kernel function $\kappa$ (\cref{eq-laplace-approx-function-space}). Contrasting \cref{eq:dual_sparse_post} and \cref{eq:gp_pred}, we can see that the computational complexity went from $\mathcal{O}(N^3)$ to $\mathcal{O}(M^3)$, with $M \ll N$.  Crucially, given the structure of our probabilistic model, our sparse dual parameters \cref{eq:dual_sparse} are a compact representation of the full model projected using the kernel.

>>>>

\begin{figure}[t!]
  % \centering\scriptsize
  \centering\Large
  % Figure options
  \pgfplotsset{axis on top,scale only axis,width=\figurewidth,height=\figureheight, ylabel near ticks,ylabel style={yshift=-2pt},y tick label style={rotate=90},legend style={nodes={scale=1., transform shape}},tick label style={font=\Large,scale=1}}
  \pgfplotsset{xlabel={Input, $x$},axis line style={rounded corners=2pt}}
  % Set figure
  % \setlength{\figurewidth}{.41\textwidth}
  \setlength{\figurewidth}{.38\textwidth}
  \setlength{\figureheight}{\figurewidth}
  %
  \def\inducing{\Huge Sparse inducing points}
  %
  \begin{subfigure}[c]{.52\columnwidth}
    \raggedleft
    \pgfplotsset{ylabel={Output, $y$}}
    \input{./fig/regression-nn.tex}%
  \end{subfigure}
  \hfill
  \begin{subfigure}[c]{.02\columnwidth}
    \centering
    \tikz[overlay,remember picture]\node(p0){};
  \end{subfigure}
  \hfill
  \begin{subfigure}[c]{.4\columnwidth}
    \raggedleft
    \pgfplotsset{yticklabels={,,},ytick={\empty}}
    \input{./fig/regression-nn2svgp.tex}%
  \end{subfigure}
%  \hfill
%  \begin{subfigure}[c]{.01\textwidth}
%    \centering
%    \tikz[overlay,remember picture]\node(p1){};
%  \end{subfigure}
%  \hfill
%  \begin{subfigure}[c]{.28\textwidth}
%    \raggedleft
%    \pgfplotsset{yticklabels={,,},ytick={\empty}}
%    \input{./fig/regression-update.tex}%
%  \end{subfigure}
  \vspace*{-0.8em}
  \caption{\textbf{Regression example on an MLP with two hidden layers.} Left:~Predictions from the trained neural network. Right:~Our approach (\our) equips trained NNs with uncertainty estimates.} %The model captures the predictive mean and uncertainty, and (right) makes it possible to incorporate new data without retraining the model.}
  % \caption{\textbf{Regression example on an MLP with two hidden layers.} Left:~Predictions from the trained neural network. Right:~Our approach (\our) summarizes all the training data with the help of a set of inducing points.} %The model captures the predictive mean and uncertainty, and (right) makes it possible to incorporate new data without retraining the model.}
  \label{fig:teaser}
  %
  \begin{tikzpicture}[remember picture,overlay]
    % Arrow style
    \tikzstyle{myarrow} = [draw=black!80, single arrow, minimum height=14mm, minimum width=2pt, single arrow head extend=4pt, fill=black!80, anchor=center, rotate=0, inner sep=5pt, rounded corners=1pt]
    % Arrows
    \node[myarrow] (p0-arr) at ($(p0) + (0.2em,1.2em)$) {};
    % \node[myarrow] (p0-arr) at ($(p0) + (1em,1.5em)$) {};
    %\node[myarrow] (p1-arr) at ($(p1) + (1em,1.5em)$) {};
    % Arrow labels
    \node[font=\scriptsize\sc,color=white] at (p0-arr) {\our};
    %\node[font=\scriptsize\sc,color=white] at (p1-arr) {new data};
  \end{tikzpicture}
  \vspace*{-1em}
\end{figure}

@@

\begin{figure*}[t]
  \centering\scriptsize
  % \setlength{\figurewidth}{.18\textwidth}
  \setlength{\figurewidth}{.27\textwidth}
  \setlength{\figureheight}{\figurewidth}
  \pgfplotsset{axis on top,scale only axis,y tick label style={rotate=90}, x tick label style={font=\footnotesize},y tick label style={font=\footnotesize},title style={yshift=-4pt,font=\large}, y label style={font=\large},x label style={font=\large},grid=major, width=\figurewidth, height=\figureheight}
  \pgfplotsset{grid style={line width=.1pt, draw=gray!10,dashed}}
  \pgfplotsset{xlabel={$M$ as \% of $N$},ylabel style={yshift=-12pt}}
  %
  \begin{minipage}[t]{.155\textwidth}
    \raggedleft
    \pgfplotsset{ylabel=NLPD}
    \input{./fig/australian.tex}
  \end{minipage}
  \hfill
  % \begin{minipage}[t]{.12\textwidth}
  %   \raggedleft
  %   \input{./fig/breast_cancer.tex}
  % \end{minipage}
  % \hfill
  % \begin{minipage}[t]{.12\textwidth}
  %   \raggedleft
  %   \input{./fig/ionosphere.tex}
  % \end{minipage}
  % \hfill
  \begin{minipage}[t]{.155\textwidth}
    \raggedleft
    \input{./fig/glass.tex}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{.155\textwidth}
    \raggedleft
    \input{./fig/vehicle.tex}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{.155\textwidth}
    \raggedleft
    \input{./fig/waveform.tex}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{.155\textwidth}
    \raggedleft
    \input{./fig/digits.tex}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{.155\textwidth}
    \raggedleft
    \input{./fig/satellite.tex}
  \end{minipage}\\[-1em]
  %
  % Legend
  \definecolor{steelblue31119180}{RGB}{31,119,180}
  \definecolor{darkorange25512714}{RGB}{255,127,14}
  \newcommand{\myline}[1]{\protect\tikz[baseline=-.5ex,line width=1.6pt]\protect\draw[draw=#1](0,0)--(1.2em,0);}
  \caption{Comparison of convergence in terms of number of inducing points $M$ in NLPD (mean\textcolor{gray}{\footnotesize$\pm$std} over 5 seeds) on UCI classification tasks: \our (\myline{steelblue31119180}) vs. GP subset (\myline{darkorange25512714}). Our \our converges fast for all cases showing clear benefits of its ability to summarize all the data onto a sparse set of inducing points. The number of inducing points $M$ are specified as a percent of the number of training data $N$.}
  \label{fig:uci}
  %\vspace*{-6pt}
\end{figure*}

@@

As demonstrated in \citet{adam2021dual}, the posterior under this model bears a structure akin to \cref{eq:gp_pred}. The authors of that paper exploit the dual parameters for approximate inference, but write them using the natural parameterization, primarily because this form is more suitable for optimization through natural gradients.
We project the dual parameters onto the inducing points giving sparse dual parameters,
% In order to link the dual parameters defined in \cref{eq:dual_param}, we write them in the sparse GP using the dual parameters,
%
\begin{equation} \textstyle
  \valpha_{\vu}  =  \sum_{i=1}^N  \vkzi \, \hat{\alpha}_{i}
  \quad \text{and} \quad
  \MBeta_{\vu} =  \sum_{i=1}^N \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,
\label{eq:dual_sparse}
\end{equation}
%
where the sparse dual parameters are now a sum over \emph{all data points}, with $\valpha_{\vu} \in \R^{M}$ and $\MBeta_{\vu} \in \R^{M  \times M}$. Using this sparse definition of the dual parameters, our sparse GP posterior takes the following form:
\begin{align}\label{eq:dual_sparse_post}
   \myexpect_{q_{\vu}(\vf)}[f_i] &= \vkzs^{\T} \MKzz^{-1} \valpha_{\vu}, \\
   %\quad \text{and} \quad
   \textrm{Var}_{q_{\vu}(\vf)}[f_i]  &= k_{ii} - \vkzs^\top [\MKzz^{-1} - (\MKzz + \MBeta_{\vu})^{-1} ]\vkzs, \nonumber
\end{align}
where $\MKzz$ and $\vkzs$ are defined similarly to $\MKxx$ and $\vk_{\vx i}$ but over the inducing points $\{\vz_j\}_{i=1}^M$, $\vz \in \R^{D}$, instead of the full data set $\dataset$. The key quantities we need to make predictions from our sparse GP from the converged NN are $(\hat{\alpha}_i, \hat{\beta}_i)$ (\cref{eq:dual_param_laplace}) and a kernel function $\kappa$ (\cref{eq-laplace-approx-function-space}). Contrasting \cref{eq:dual_sparse_post} and \cref{eq:gp_pred}, we can see that the computational complexity went from $\mathcal{O}(N^3)$ to $\mathcal{O}(M^3)$, with $M \ll N$.  Crucially, given the structure of our probabilistic model, our sparse dual parameters \cref{eq:dual_sparse} are a compact representation of the full model projected using the kernel.
